Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities

### 2019-06-19 paper.Rmd
major update

---
title: Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities
date: ''
output: 
  pdf_document:
    keep_tex: true
    includes:
      in_header: preamble.sty
fig_width: 4
fig_height: 2.6
bibliography: library.bib
---

@@

Mozambique's economic growth up to 2014 was facilitated by the government's open policy towards foreign direct investment (FDI), a plentiful supply of natural resources [@Rogers], and relatively inexpensive labor. Foreign investment in Mozambique is largely geared towards the export market, and heavily concentrated. 63% of exports come from aluminum, electricity, minerals, and gas, with each sector displaying high degrees of concentration (greater than 50% of the export share of the previous four industries are attributable to one company) [@Sutton2014]. The total number of foreign enterprises operating in Mozambique is not ascertainable, but in both quantity and diversity of sources, FDI has increased dramatically in recent years [@Sutton2014]. Mozambique was the Sub-Saharan African country with the greatest increase in FDI (defined by the World Bank as cross-border investment to establish a lasting interest) inflows from 2006 to 2014, registering a boom between 2010 and 2014 of about 5000 million USD [@Unctad2012].

@@

What stands out is the presence of many "non-Western" countries in the top 10. Though reliable data only go as far back as 2009, the landscape of investment 20 years prior would have been both (a) far smaller and (b) primarily composed of western countries, especially given that Mozambique emerged so late from civil war (1994). Though as of 2015, China did not figure in the top 10 FDI countries in Mozambique, its upward trend is notable, with approximately 1.5 million USD invested in 2010 to nearly 100 million by 2015 (a 60-fold increase).

@@

# Clean up columns
db <- data.frame(db)
macro_names <- names(db)
macro_names <- ifelse(grepl('X__', macro_names), 'Var', macro_names)

@@

# Get columns for coordinates
temp$x <- coordinates(temp)[,1]
temp$y <- coordinates(temp)[,2]

@@

# Bring in gdp data to fortified map
temp_fort <- left_join(x = temp_fort,
                       y = temp@data %>%
                         mutate(id = as.character(ID_1)))

@@

# # Convert raster to polygons
# rtp <- rasterToPolygons(r)
# rtp@data$id <- 1:nrow(rtp@data) 
```

@@

Firms with CSR activities are often large and foreign.  Among the largest "key players" in Mozambican CSR are Coca-Cola, British Petroleum and Colgate-Palmolive.  State conglomerates, such as Águas de Moçambique and Electricidade de Moçambique, also participate in CSR activities [@Compact2007]. Large communications firms, such as Vodacom and MCEL, and banks (BIM and BCI) have a CSR component, but generally focus on culture, environment and sport, avoiding activities which could either complement or conflict with the health sector [@Kaufmann_2015].

@@

The intensity of CSR activity seems to closely track FDI, but generally avoids malaria control. 
The few malaria-related CSR projects in recent years has come in large, "mega-projects" whose well-publicized malaria abatement activities are profit-driven [@Mouzin2011] or whose primary aim is not profit-related [@Han]. Though the former is often portrayed as a "win-win" for business and public health, the latter also offers tangible benefits for private industry, and should be understood as operating under the same conditions and with the same motivations. To the extent that most CSR activity has avoided the health sector (including malaria) in favor of cultural, environmental, education and more general "development" activities, it is reasonable to assume that this may be due to the perceived costs of malaria control, lack of perceived PR benefits, the government and NGO's predominant role in the area, as well as the "opt-in" nature and privacy/legal issues generally related to engaging in health-related campaigns.

@@

A secondary risk is that increased private sector involvement in malaria control could cause a decrease in public sector _competence_ in the prevention and treatment of malaria. This could have negative consequences in the case of either a financial or economic crisis (in which CSR activity would be curtailed) or an increase in malarial activity. By the same token, CSR involvement in malaria control could potentially portend, at least initially, less effective interventions. Private firms' incentives, though aligned with the public's in terms of malaria, are not identical, and pressures from shareholders and for positive public perception might motivate malaria control strategies which do not necessarily carry with them the most recent scientific knowledge.

@@

High FDI in Mozambique and growing interest in CSR both call for increased reflection on the private sector's role in the delivery of public goods. A refocusing of CSR expenditures into areas where the need is greatest (specifically, malaria control) could lead to better health and greater profits. This "win-win" for the public and private sectors represents a rare opportunity, which deserves more discourse, research and experimentation.

>>>>

---
title: Foreign direct investment, corporate social responsibility, and malaria control in Mozambique - trends, risks, and opportunities
date: ''
output: 
  pdf_document:
    keep_tex: true
    includes:
      in_header: preamble.sty
fig_width: 4
fig_height: 2.6
bibliography: fdi_moz.bib
---

@@

Mozambique's economic growth up to 2014 was facilitated by the government's open policy towards foreign direct investment (FDI), a plentiful supply of natural resources [@Rogers], and relatively inexpensive labor. These factors combined to push Maputo (Mozambique's capital) into the top-20 African countries in terms of gross FDI [@africa2018]. Its location (a large, eastward-facing coastline, proximity to rapidly developing South Africa, and integration into trade corridor and zones such as those with Zimbabwe [@africa2018] and the Southern African community as a whole [@Asiedu_2005]) are favorable to foreign investment, which is largely geared towards the export market, and heavily concentrated. 63% of exports come from aluminum, electricity, minerals, and gas, with each sector displaying high degrees of concentration (greater than 50% of the export share of the previous four industries are attributable to one company) [@Sutton2014]. The total number of foreign enterprises operating in Mozambique is not ascertainable, but in both quantity and diversity of sources, FDI has increased dramatically in recent years [@Sutton2014]. Mozambique was the Sub-Saharan African country with the greatest increase in FDI (defined by the World Bank as cross-border investment to establish a lasting interest) inflows from 2006 to 2014, registering a boom between 2010 and 2014 of about 5000 million USD [@Unctad2012].

@@

What stands out is the presence of many "non-Western" countries in the top 10. Though reliable data only go as far back as 2009, the landscape of investment 20 years prior would have been both (a) far smaller and (b) primarily composed of western countries, especially given that Mozambique emerged so late from civil war (1994). As of 2014, Chinese capital had entered into 52 countries in Africa [@africa2018], and though more recent data are not available, it's likely that this figure has grown since then. Though as of 2015, China did not figure in the top 10 FDI countries in Mozambique, its upward trend is notable, with approximately 1.5 million USD invested in 2010 to nearly 100 million by 2015 (a 60-fold increase). As of 2017, 3.2% of Chinese FDI flows went to Mozambique [@He_2015]. Though much of the growth is fueled by centralized State actors, an increasing number of provincial and decentralized investors are playing a role in African FDI [@Gu_2016].

@@

# Clean up columns
db <- data.frame(db)
macro_names <- names(db)
macro_names <- gsub('...', 'X__...', macro_names, fixed = TRUE)
macro_names <- ifelse(grepl('X__', macro_names), 'Var', macro_names)

@@

# Get columns for coordinates
temp@data$x <- coordinates(temp)[,1]
temp@data$y <- coordinates(temp)[,2]

# Get a better id
temp@data$id <-
  unlist(lapply(strsplit(gsub('MOZ.', '', temp@data$GID_1), '_'), function(x){x[1]}))

@@

# Bring in gdp data to fortified map
temp_fort <- left_join(x = temp_fort,
                       y = temp@data)

@@

# # Convert raster to polygons
# rtp <- rasterToPolygons(r)
# rtp@data$id <- 1:nrow(rtp@data)

@@

Firms with CSR activities are often large and foreign.  Among the largest "key players" in Mozambican CSR are Coca-Cola, British Petroleum and Colgate-Palmolive.  State conglomerates, such as Águas de Moçambique and Electricidade de Moçambique, also participate in CSR activities [@Compact2007]. Large communications firms, such as Vodacom and MCEL, and banks (BIM and BCI) have a CSR component, but generally focus on culture, environment and sport, avoiding activities which could either complement or conflict with the health sector. Much of CSR in Mozambique is basic servcies only, a phenomenon Kaufmann described as "CSR towards compliance" as opposed to "beyond compliance" [@Kaufmann_2015].

@@

The intensity of CSR activity seems to closely track FDI, but generally avoids malaria control. 
The few malaria-related CSR projects in recent years has come in large, "mega-projects" whose well-publicized malaria abatement activities are profit-driven [@Mouzin2011] or whose primary aim is not profit-related [@Han]. This should come as no surprise given that "mega-projects" make up a majority of FDI to Mozambique (albeit declining slightly in recent years) [@Unctad2012]. Though the former is often portrayed as a "win-win" for business and public health, the latter also offers tangible benefits for private industry, and should be understood as operating under the same conditions and with the same motivations. To the extent that most CSR activity has avoided the health sector (including malaria) in favor of cultural, environmental, education and more general "development" activities, it is reasonable to assume that this may be due to the perceived costs of malaria control, lack of perceived PR benefits, the government and NGO's predominant role in the area, as well as the "opt-in" nature and privacy/legal issues generally related to engaging in health-related campaigns.

@@

A secondary risk is that increased private sector involvement in malaria control could cause a decrease in public sector _competence_ in the prevention and treatment of malaria. This could have negative consequences in the case of either a financial or economic crisis (in which CSR activity would be curtailed) or an increase in malarial activity. By the same token, CSR involvement in malaria control could potentially portend, at least initially, less effective interventions. Private firms' incentives, though aligned with the public's in terms of malaria, are not identical, and pressures from shareholders and for positive public perception might motivate malaria control strategies which do not necessarily carry with them the most recent scientific knowledge. The methods by which private initiatives could contribute to malaria elimination and control while at the same time only "complementing" (rather than "substituting") State activities is not entirely clear, and would require significant organization [@Kaufmann_2015].

@@

High FDI in Mozambique and growing interest in CSR both call for increased reflection on the private sector's role in the delivery of public goods. A refocusing of CSR expenditures into areas where the need is greatest (specifically, malaria control) could lead to better health and greater profits, and go beyond the perception of superficial CSR which doesn't address core social problems (ie, "green washing") [@Kaufmann_2015]. This "win-win" for the public and private sectors represents a rare opportunity, which deserves more discourse, research and experimentation.

### 2019-06-19 paper.tex
major update

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Foreign direct investment, corporate social responsibility, and malaria
control in Mozambique - trends, risks, and opportunities}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

Mozambique's economic growth up to 2014 was facilitated by the
government's open policy towards foreign direct investment (FDI), a
plentiful supply of natural resources (Rogers 2014), and relatively
inexpensive labor. Foreign investment in Mozambique is largely geared
towards the export market, and heavily concentrated. 63\% of exports
come from aluminum, electricity, minerals, and gas, with each sector
displaying high degrees of concentration (greater than 50\% of the
export share of the previous four industries are attributable to one
company) (Sutton 2014). The total number of foreign enterprises
operating in Mozambique is not ascertainable, but in both quantity and
diversity of sources, FDI has increased dramatically in recent years
(Sutton 2014). Mozambique was the Sub-Saharan African country with the
greatest increase in FDI (defined by the World Bank as cross-border
investment to establish a lasting interest) inflows from 2006 to 2014,
registering a boom between 2010 and 2014 of about 5000 million USD
(UNCTAD 2012).

@@

Motivations for engaging in CSR has been analyzed in the economic
literature. Enterprises would have no incentives to engage in CSR
according to Friedman (Friedman 1970) as firms are organized to maximize
their profits only. The opposite view is that CSR is actually a strategy
for maximizing profits. According to this vision, CSR helps enterprises
to make their products known and, therefore, foment consumers' demand
(Porter and Kramer 2002). CSR also helps building a positive reputation
of enterprises among institutions and communities where they operate:
this is particularly useful for foreign firms that have no previous
contacts with the host country/communities (Zhu, Sun, and Leung 2013).

@@

What stands out is the presence of many ``non-Western'' countries in the
top 10. Though reliable data only go as far back as 2009, the landscape
of investment 20 years prior would have been both (a) far smaller and
(b) primarily composed of western countries, especially given that
Mozambique emerged so late from civil war (1994). Though as of 2015,
China did not figure in the top 10 FDI countries in Mozambique, its
upward trend is notable, with approximately 1.5 million USD invested in
2010 to nearly 100 million by 2015 (a 60-fold increase).

@@

Firms with CSR activities are often large and foreign. Among the largest
``key players'' in Mozambican CSR are Coca-Cola, British Petroleum and
Colgate-Palmolive. State conglomerates, such as Águas de Moçambique and
Electricidade de Moçambique, also participate in CSR activities (Compact
2007). Large communications firms, such as Vodacom and MCEL, and banks
(BIM and BCI) have a CSR component, but generally focus on culture,
environment and sport, avoiding activities which could either complement
or conflict with the health sector (Kaufmann and Simons-Kaufmann 2015).

@@

The intensity of CSR activity seems to closely track FDI, but generally
avoids malaria control. The few malaria-related CSR projects in recent
years has come in large, ``mega-projects'' whose well-publicized malaria
abatement activities are profit-driven (Mouzin and al. 2011) or whose
primary aim is not profit-related (Han 2015). Though the former is often
portrayed as a ``win-win'' for business and public health, the latter
also offers tangible benefits for private industry, and should be
understood as operating under the same conditions and with the same
motivations. To the extent that most CSR activity has avoided the health
sector (including malaria) in favor of cultural, environmental,
education and more general ``development'' activities, it is reasonable
to assume that this may be due to the perceived costs of malaria
control, lack of perceived PR benefits, the government and NGO's
predominant role in the area, as well as the ``opt-in'' nature and
privacy/legal issues generally related to engaging in health-related
campaigns.

@@

A secondary risk is that increased private sector involvement in malaria
control could cause a decrease in public sector \emph{competence} in the
prevention and treatment of malaria. This could have negative
consequences in the case of either a financial or economic crisis (in
which CSR activity would be curtailed) or an increase in malarial
activity. By the same token, CSR involvement in malaria control could
potentially portend, at least initially, less effective interventions.
Private firms' incentives, though aligned with the public's in terms of
malaria, are not identical, and pressures from shareholders and for
positive public perception might motivate malaria control strategies
which do not necessarily carry with them the most recent scientific
knowledge.

@@

High FDI in Mozambique and growing interest in CSR both call for
increased reflection on the private sector's role in the delivery of
public goods. A refocusing of CSR expenditures into areas where the need
is greatest (specifically, malaria control) could lead to better health
and greater profits. This ``win-win'' for the public and private sectors
represents a rare opportunity, which deserves more discourse, research
and experimentation.

@@

@@

@@

@@

>>>>

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Foreign direct investment, corporate social responsibility, and malaria
control in Mozambique - trends, risks, and opportunities}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Mozambique's economic growth up to 2014 was facilitated by the
government's open policy towards foreign direct investment (FDI), a
plentiful supply of natural resources (Rogers 2014), and relatively
inexpensive labor. These factors combined to push Maputo (Mozambique's
capital) into the top-20 African countries in terms of gross FDI
(UNHabitat 2018). Its location (a large, eastward-facing coastline,
proximity to rapidly developing South Africa, and integration into trade
corridor and zones such as those with Zimbabwe (UNHabitat 2018) and the
Southern African community as a whole (Asiedu 2005)) are favorable to
foreign investment, which is largely geared towards the export market,
and heavily concentrated. 63\% of exports come from aluminum,
electricity, minerals, and gas, with each sector displaying high degrees
of concentration (greater than 50\% of the export share of the previous
four industries are attributable to one company) (Sutton 2014). The
total number of foreign enterprises operating in Mozambique is not
ascertainable, but in both quantity and diversity of sources, FDI has
increased dramatically in recent years (Sutton 2014). Mozambique was the
Sub-Saharan African country with the greatest increase in FDI (defined
by the World Bank as cross-border investment to establish a lasting
interest) inflows from 2006 to 2014, registering a boom between 2010 and
2014 of about 5000 million USD (UNCTAD 2012).

@@

Motivations for engaging in CSR has been analyzed in the economic
literature. Enterprises would have no incentives to engage in CSR
according to Friedman (Friedman 1970) as firms are organized to maximize
their profits only. The opposite view is that CSR is actually a strategy
for maximizing profits. According to this vision, CSR helps enterprises
to make their products known and, therefore, foment consumers' demand
(Porter and Kramer 2002). CSR also helps building a positive reputation
of enterprises among institutions and communities where they operate:
this is particularly useful for foreign firms that have no previous
contacts with the host country/communities (Y. Zhu, Sun, and Leung
2013).

@@

What stands out is the presence of many ``non-Western'' countries in the
top 10. Though reliable data only go as far back as 2009, the landscape
of investment 20 years prior would have been both (a) far smaller and
(b) primarily composed of western countries, especially given that
Mozambique emerged so late from civil war (1994). As of 2014, Chinese
capital had entered into 52 countries in Africa (UNHabitat 2018), and
though more recent data are not available, it's likely that this figure
has grown since then. Though as of 2015, China did not figure in the top
10 FDI countries in Mozambique, its upward trend is notable, with
approximately 1.5 million USD invested in 2010 to nearly 100 million by
2015 (a 60-fold increase). As of 2017, 3.2\% of Chinese FDI flows went
to Mozambique (He, Xie, and Zhu 2015). Though much of the growth is
fueled by centralized State actors, an increasing number of provincial
and decentralized investors are playing a role in African FDI (Gu et al.
2016).

@@

Firms with CSR activities are often large and foreign. Among the largest
``key players'' in Mozambican CSR are Coca-Cola, British Petroleum and
Colgate-Palmolive. State conglomerates, such as Águas de Moçambique and
Electricidade de Moçambique, also participate in CSR activities (Compact
2007). Large communications firms, such as Vodacom and MCEL, and banks
(BIM and BCI) have a CSR component, but generally focus on culture,
environment and sport, avoiding activities which could either complement
or conflict with the health sector. Much of CSR in Mozambique is basic
servcies only, a phenomenon Kaufmann described as ``CSR towards
compliance'' as opposed to ``beyond compliance'' (Kaufmann and
Simons-Kaufmann 2015).

@@

The intensity of CSR activity seems to closely track FDI, but generally
avoids malaria control. The few malaria-related CSR projects in recent
years has come in large, ``mega-projects'' whose well-publicized malaria
abatement activities are profit-driven (Mouzin and al. 2011) or whose
primary aim is not profit-related (Han 2015). This should come as no
surprise given that ``mega-projects'' make up a majority of FDI to
Mozambique (albeit declining slightly in recent years) (UNCTAD 2012).
Though the former is often portrayed as a ``win-win'' for business and
public health, the latter also offers tangible benefits for private
industry, and should be understood as operating under the same
conditions and with the same motivations. To the extent that most CSR
activity has avoided the health sector (including malaria) in favor of
cultural, environmental, education and more general ``development''
activities, it is reasonable to assume that this may be due to the
perceived costs of malaria control, lack of perceived PR benefits, the
government and NGO's predominant role in the area, as well as the
``opt-in'' nature and privacy/legal issues generally related to engaging
in health-related campaigns.

@@

A secondary risk is that increased private sector involvement in malaria
control could cause a decrease in public sector \emph{competence} in the
prevention and treatment of malaria. This could have negative
consequences in the case of either a financial or economic crisis (in
which CSR activity would be curtailed) or an increase in malarial
activity. By the same token, CSR involvement in malaria control could
potentially portend, at least initially, less effective interventions.
Private firms' incentives, though aligned with the public's in terms of
malaria, are not identical, and pressures from shareholders and for
positive public perception might motivate malaria control strategies
which do not necessarily carry with them the most recent scientific
knowledge. The methods by which private initiatives could contribute to
malaria elimination and control while at the same time only
``complementing'' (rather than ``substituting'') State activities is not
entirely clear, and would require significant organization (Kaufmann and
Simons-Kaufmann 2015).

@@

High FDI in Mozambique and growing interest in CSR both call for
increased reflection on the private sector's role in the delivery of
public goods. A refocusing of CSR expenditures into areas where the need
is greatest (specifically, malaria control) could lead to better health
and greater profits, and go beyond the perception of superficial CSR
which doesn't address core social problems (ie, ``green washing'')
(Kaufmann and Simons-Kaufmann 2015). This ``win-win'' for the public and
private sectors represents a rare opportunity, which deserves more
discourse, research and experimentation.

@@

\hypertarget{ref-Asiedu_2005}{}
Asiedu, Elizabeth. 2005. ``Foreign Direct Investment in Africa: The Role
of Natural Resources, Market Size, Government Policy, Institutions and
Political Instability.'' \emph{SSRN Electronic Journal}. Elsevier BV.
doi:\href{https://doi.org/10.2139/ssrn.717361}{10.2139/ssrn.717361}.

@@

\hypertarget{ref-Gu_2016}{}
Gu, Jing, Chuanhong Zhang, Alcides Vaz, and Langton Mukwereza. 2016.
``Chinese State Capitalism? Rethinking the Role of the State and
Business in Chinese Development Cooperation in Africa.'' \emph{World
Development} 81 (May). Elsevier BV: 24--34.
doi:\href{https://doi.org/10.1016/j.worlddev.2016.01.001}{10.1016/j.worlddev.2016.01.001}.

@@

\hypertarget{ref-He_2015}{}
He, Canfei, Xiuzhen Xie, and Shengjun Zhu. 2015. ``Going Global:
Understanding China's Outward Foreign Direct Investment from
Motivational and Institutional Perspectives.'' \emph{Post-Communist
Economies} 27 (4). Informa UK Limited: 448--71.
doi:\href{https://doi.org/10.1080/14631377.2015.1084716}{10.1080/14631377.2015.1084716}.

@@

\hypertarget{ref-africa2018}{}
UNHabitat. 2018. \emph{The State of African Cities 2018, The geography
of African investment}. United Nations.
